---
title: 'ESM 244: Lab 1'
author: 'By: Claire Gonzales'
date: "1/5/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message=FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)

```

### Reading in the Data

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"))

```

### Part 1: wrangling and ggplot review

**Example 1: ** Find counts of observation by `legal_status` & wrangle a bit

```{r}
### method 1: group_by() %>% summarize()
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

### method 2: different way plus a few new functions
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% # will bring this column to the front
  slice_max(tree_count, n = 5) %>%  # slices at top 5
  arrange(desc(tree_count))

```

Make a graph of the top 5

```{r}
# using fct_reorder to put cols in desc order of tree_count
ggplot(data = top_5_status, aes(x = fct_reorder (legal_status, tree_count), y = tree_count)) +
  geom_col(fill = 'darkgreen') +
  labs(x = "Legal Status",
       y = "Tree Count") +
  coord_flip() +
  theme_minimal()
```

**Example 2** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA", and store as 'permitted_data_df'

```{r}
# sf_trees$legal_status %>%  unique()
permitted_data_df <- sf_trees %>% 
  filter(legal_status == "Permitted Site", caretaker == "MTA")
```

**Example 3:** only keep Blackwood Acacia trees and then only keep columns `legal_status`, `data`, `latitude` and store as `blackwood_acacia_df`:

```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% 
  select(legal_status, date, lat = latitude, lon = longitude)

### Make a graph of this
ggplot(data = blackwood_acacia_df, aes(x = lon, y= lat)) +
  geom_point(color = "darkgreen")
```

**Example 4:** Use `tidy::separate()`

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c("spp_scientific", "spp_common"), sep = " :: ")
```

**Example 5:** Use `tidy::unite()`

```{r}
ex_5 <- sf_trees %>% 
  unite("id_status", tree_id, legal_status, sep = "_COOL_")
```

### Part 2: make some maps

**Step 1:** convert the lat/lon to spatial point, `st_as_sf()`

```{r}
# this function will take lat and long and convert to spatial points

blackwood_acaia_sf <- blackwood_acacia_df %>% 
  drop_na(lat, lon) %>% 
  st_as_sf(coords = c("lon", "lat"))

# we need to tell r what the coord reference system is using `st_crs`
st_crs(blackwood_acacia_sf) <- 4326

ggplot(data = blackwood_acaia_sf) +
  geom_sf(color = "darkgreen")+
  theme_minimal()
```

Read in the SF shapefile and add to map

```{r}

sf_map <- read_sf(here("data", "sf_map", "tl_2017_06075_roads.shp"))

sf_map_transform <- st_transform(sf_map, 4326)

ggplot(data = sf_map_transform) +
  geom_sf()+
  theme_minimal()
```

Combine the maps!

```{r}
ggplot()+
  geom_sf(data = sf_map,
          size = .1,
          color = "darkgrey") +
  geom_sf(data = blackwood_acacia_sf,
          color = "red",
          size = 0.5) +
  theme_void()+
  labs(title = "Blackwood acacias in SF")
```

### Now an interactive map

```{r}
tmap_mode("view")

tm_shape(blackwood_acacia_sf) +
  tm_dots()
```


